Global optimization method with dual Lipschitz constant estimates for problems with non-convex constraints

### 2020-01-24 2019/NUMTA_Barkalov/Special Issure/article.tex
Added Numerical experiments

\section{Numerical experiments}
\begin{figure}[h!]
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Simple_59_IA} \\ a)}
	\end{minipage}
	\hfill
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Simple_59_IA-DL} \\ b)}
	\end{minipage}
	\caption{Solving a Simple problem by IA (a) and IA-DL (b) methods}
	\label{fig:examples_simple}
\end{figure}

\begin{figure}[h!]
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Hard_7_IA} \\ a)}
	\end{minipage}
	\hfill
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Hard_7_IA-DL} \\ b)}
	\end{minipage}
	\caption{Solving a Hard problem by IA (a) and IA-DL (b) methods}
	\label{fig:examples_simple}
\end{figure}

\begin{table}[h!]
	\centering
		\begin{tabular}{ccccccccccccc}
			\hline
			& \multicolumn{2}{c}{$N=2$} & & \multicolumn{2}{c}{$N=3$} & & \multicolumn{2}{c}{$N=4$} & & \multicolumn{2}{c}{$N=5$} \\
			\cline{2-3} \cline{5-6} \cline{8-9} \cline{11-12}
			& \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard} \\
			\hline
			IA-DL, $r_{\nu}^{loc}=1.4$ & 316 & 766 & & 7403 & 9636 & & 44304 & 60904 & & 296357 & 763711 \\
			\hline
			IA-DL, $r_{\nu}^{loc}=2.0$ & 350 & 830 & & 8990(1) & 12020 & & 47133 & 57473 & & 360497 & 743874(1) \\
			\hline
			IA & 439 & 1074 & & 11804(1) & 20327 & & 42837 & 87310(2) & & 404660(2) & 1084990(2) \\
			\hline
		\end{tabular}
	\caption{Average number of iterations}
	\label{tab:average_number_of_iterations}
\end{table}

\begin{table}[h!]
	\centering
		\begin{tabular}{ccccc}
			\hline
			& $N=2$ & $N=3$ & $N=4$ & $N=5$ \\
			\hline
			IA-DL, $r_{\nu}^{loc}=1.4$ & 316 & 7403 & 44304 & 296357 \\
			\hline
			IA-DL, $r_{\nu}^{loc}=2.0$ & 350 & 8990(1) & 47133 & 360497 \\
			\hline
			IA & 439 & 11804(1) & 42837 & 404660(2) \\
			\hline
		\end{tabular}
	\caption{Average number of iterations Simple}
	\label{tab:average_number_of_iterations_simple}
\end{table}

\begin{table}[h!]
	\centering
		\begin{tabular}{ccccc}
			\hline
			& $N=2$ & $N=3$ & $N=4$ & $N=5$ \\
			\hline
			IA-DL, $r_{\nu}^{loc}=1.4$ & 766 & 9636 & 60904 & 763711 \\
			\hline
			IA-DL, $r_{\nu}^{loc}=2.0$ & 830 & 12020 & 57473 & 743874(1) \\
			\hline
			IA & 1074 & 20327 & 87310(2) & 1084990(2) \\
			\hline
		\end{tabular}
	\caption{Average number of iterations Hard}
	\label{tab:average_number_of_iterations_hard}
\end{table}

\begin{figure}[h!]
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Operating_characteristic_gklssC_3} \\ a)}
	\end{minipage}
	\hfill
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Operating_characteristic_gklshC_3} \\ b)}
	\end{minipage}
	\caption{Operational characteristics for Simple (a) and Hard (b) problems, $N=3$}
	\label{fig:operational_characteristics_N=3}
\end{figure}

\begin{figure}[h!]
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Operating_characteristic_gklssC_5} \\ a)}
	\end{minipage}
	\hfill
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Operating_characteristic_gklshC_5} \\ b)}
	\end{minipage}
	\caption{Operational characteristics for Simple (a) and Hard (b) problems, $N=5$}
	\label{fig:operational_characteristics_N=5}
\end{figure}

\begin{figure}[h!]
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Decision_rule_gklssC_4} \\ a)}
	\end{minipage}
	\hfill
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Decision_rule_gklshC_4} \\ b)}
	\end{minipage}
	\caption{Frequency of using various estimates of the Lipschitz constant for Simple (a) and Hard (b) problems, $N=4$}
	\label{fig:frequency_N=4}
\end{figure}

@@

\section{TEST}
\cite{Evtushenko1971, Piyavskii1972, Shubert1972, Strongin1970, Evtushenko2009, Evtushenko2013, Strongin2000, Sergeyev2013, Pinter1996, Jones2009, Wood1991, Meewella1988, Mladineo1986, Vaz2009, Stripinis2019, Paulavicius2016, Pillo2012, Pillo2016, Barkalov2017_1, Barkalov2017_2, Sergeyev2006, Zilinskas2008, Sovrasov2019, Kvasov2003, Sergeyev2010,Sergeyev2016, Horst1996, Gablonsky2001, Jones1993, Gaviano2003, Barkalov2018, Paulavicius2014, Sergeyev2015, Strongin2018, Gergel2017_2, Gergel2019}.

>>>>

\section{Numerical experiments}
	To demonstrate the efficiency of the proposed algorithm, let us consider the results of solving several series of multidimensional constrained global optimization problems. In order to construct the problems, the functions generated by the GKLS generator \cite{Gaviano2003} were used. This generator of multiextremal functions is often used for investigating global optimization algorithms \cite{Barkalov2018, Paulavicius2014, Sergeyev2015}. According to \cite{Gaviano2003}, we used simpler and more complex functions, which are  further denoted as “Simple” and “Hard”, respectively. The complexity of the functions is determined, in particular, by the radius of the global minimizer's attraction region and by the distance between global and local minimizers.

	In \cite{Gergel2017_2, Gergel2019}, an approach was proposed that allows constructing the constrained global optimization problems using certain classes of multiextremal functions (in particular, the GKLS functions), in which:
\begin{itemize}
  \item the number of constraints can be varied;
  \item the volume of the feasible subdomain $Q_{m+1}$ relative to the whole search domain $D$ is defined;
	\item The global minimizer is known in advance and can be located at the boundary of the feasible domain;
	\item The number of constraints active in the global minimizer can be predefined.
\end{itemize}

	In the experiments that we have conducted, the number of constraints in the problems was equal to 2, and one of these constraints was active at the point of the global minimum. For example, the contour plots of the objective functions and constraints for $N=2$ are presented in Figs. \ref{fig:examples_simple} and \ref{fig:examples_hard}; the feasible domains are highlighted by green. It can be seen that the feasible domains in these problems are non-convex and non-connected, and the global minimizers are located at the boundaries of the feasible domains. Also, one can see that the problems of the "Hard" class have, in general, a more complex structure: the global minimizers are located at the boundaries of the feasible domains of smaller areas whereas in the problems of the “Simple” class the optimizers are located at the boundaries of the subdomains of greater areas.

\begin{figure}[h!]
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Simple_59_IA} \\ a)}
	\end{minipage}
	\hfill
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Simple_59_IA-DL} \\ b)}
	\end{minipage}
	\caption{Solving a problem of the “Simple” class by IA (a) and IA-DL (b) methods}
	\label{fig:examples_simple}
\end{figure}

\begin{figure}[h!]
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Hard_7_IA} \\ a)}
	\end{minipage}
	\hfill
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Hard_7_IA-DL} \\ b)}
	\end{minipage}
	\caption{Solving a problem of the “Hard” class by IA (a) and IA-DL (b) methods}
	\label{fig:examples_hard}
\end{figure}

	Also, the points of the trials performed by (a) the original index algorithm (IA) and by (b) the method with dual estimate of the Lipschitz constant (IA-DL) are marked in Figs. \ref{fig:examples_simple} and \ref{fig:examples_hard}. To solve the problems shown in Figs. \ref{fig:examples_simple} and \ref{fig:examples_hard}, the IA method required 643 and 1166 iterations whereas the IA-DL method solved similar problems for 373 and 712 iterations, respectively.

	The reduction of the number of trials was also confirmed by solving several series of  problems composed of the functions of the “Simple” and “Hard” classes. Each series consisted of 100 problems and was solved by the IA and IA-DL methods. The global minimizer $y^{\ast}$ was considered to be found, if the algorithm generated a trial point $y^k$ in the vicinity of the global minimum i. e. $\| y^{\ast} - y^k \| < \varepsilon \|b - a\|$ where $a$ and $b$ are the boundaries of the search domain $D$. To reduce the dimensionality, the evolvent constructed with the parameter $m=10$ was used in both algorithms. The relative accuracy of the search for the solution at $N = 2, 3$ was $\epsilon=0.01$ and for the solution at $N=4,5$  it was $\epsilon=0.03$. The reservation parameters $\epsilon_{\nu}=\mu_{\nu} \delta,1 \leq \nu \leq 2$ were set according to formula (\ref{epsilon_nu}) with $\delta=0.01$. The maximum number of iterations at $N \leq 4$ was set as $K_{max} = 10^6$, and at $N=5$ as $K_{max}=7 \cdot 10^6$.

	Since all functions involved in the statement of each particular problem belong to the same class, equal parameters $r_{\nu}=4.1$ were set for the problems of the “Simple” class and $r_{\nu}=6.2$ for the problems of the “Hard” class. When solving the problems of these classes by the IA-DL method, the values of the parameters $r_{\nu}$ given above were selected for the upper estimate of the Lipschitz constants i. e. the values $r_{\nu}^{glob}=4.1$ and $r_{\nu}^{glob}=6.2$ were set, which were supplemented by various values of $r_{\nu}^{loc}$. The average number of iterations performed by the algorithms with different values of $r_{\nu}^{loc}$ is given in Table \ref{tab:average_number_of_iterations_simple} and \ref{tab:average_number_of_iterations_hard}. The number of unsolved problems is given in brackets.

\begin{table}[h!]
	\centering
		\begin{tabular}{ccccc}
			\hline
			& $N=2$ & $N=3$ & $N=4$ & $N=5$ \\
			\hline
			IA-DL, $r_{\nu}^{loc}=1.4$ & 316 & 7403 & 44304 & 296357 \\
			\hline
			IA-DL, $r_{\nu}^{loc}=2.0$ & 350 & 8990(1) & 47133 & 360497 \\
			\hline
			IA & 439 & 11804(1) & 42837 & 404660(2) \\
			\hline
		\end{tabular}
	\caption{Average number of iterations performed by the IA and IA-DL methods on “Simple” class}
	\label{tab:average_number_of_iterations_simple}
\end{table}

\begin{table}[h!]
	\centering
		\begin{tabular}{ccccc}
			\hline
			& $N=2$ & $N=3$ & $N=4$ & $N=5$ \\
			\hline
			IA-DL, $r_{\nu}^{loc}=1.4$ & 766 & 9636 & 60904 & 763711 \\
			\hline
			IA-DL, $r_{\nu}^{loc}=2.0$ & 830 & 12020 & 57473 & 743874(1) \\
			\hline
			IA & 1074 & 20327 & 87310(2) & 1084990(2) \\
			\hline
		\end{tabular}
	\caption{Average number of iterations performed by the IA and IA-DL methods on “Hard” class}
	\label{tab:average_number_of_iterations_hard}
\end{table}

	In addition to the average number of trials required to solve the problems, the results of solving the series of problems can be also presented by the function $p(k)$, which describes the number of problems solved for $k$ iterations. This function will be called the \textit{operational characteristic} of the algorithms.

	The operational characteristics for IA and IA-DL obtained by solving the series of problems of the “Simple” and “Hard” classes with the dimensionalities $N=3$ and $N=5$ are presented in Figs. \ref{fig:operational_characteristics_N=3} and \ref{fig:operational_characteristics_N=5}, respectively. The values of parameters $r_{\nu}$ used to evaluate the Lipschitz constant are given in the figures.

\begin{figure}[h!]
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Operating_characteristic_gklssC_3} \\ a)}
	\end{minipage}
	\hfill
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Operating_characteristic_gklshC_3} \\ b)}
	\end{minipage}
	\caption{Operational characteristics for “Simple” (a) and “Hard” (b) problem classes, $N = 3$}
	\label{fig:operational_characteristics_N=3}
\end{figure}

\begin{figure}[h!]
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Operating_characteristic_gklssC_5} \\ a)}
	\end{minipage}
	\hfill
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Operating_characteristic_gklshC_5} \\ b)}
	\end{minipage}
	\caption{Operational characteristics for “Simple” (a) and “Hard” (b) problem classes, $N = 5$}
	\label{fig:operational_characteristics_N=5}
\end{figure}

	The lower (solid) curves in Figs. \ref{fig:operational_characteristics_N=3} and \ref{fig:operational_characteristics_N=5} characterize IA whereas the upper (dashed) lines characterize IA-DL. This arrangement of curves shows that the algorithm with dual estimates of the Lipschitz constant  provides, on average, a faster solution to the problems of the series than the algorithm using a single estimate of the Lipschitz constant.

	The frequency of the use of the upper and lower estimates of the Lipschitz constant in the rules of the algorithm that corresponds to the choice of $r_{\nu}=r_{\nu}^{loc}$ or $r_{\nu}=r_{\nu}^{glob}$ according to Step 6 of the algorithm is an interesting indicator. This frequency can be characterized by the average values $K_{glob}$ and $K_{loc}$ of the use of the respective (global or local) characteristic. In order to present this frequency, let us plot the number of trials normalized from 1 to 100 on the abscissa axis. Actually, in this case the abscissa axis will reflect the progress in solving the problem (in percentage) where 1 corresponds to the beginning and 100 -- to the end of the process of the search for the solution. The indicators $K_{glob}$ and $K_{loc}$ will be plotted on the ordinate axis. Such plots are presented in Fig. \ref{fig:frequency_N=4}.

\begin{figure}[h!]
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Decision_rule_gklssC_4} \\ a)}
	\end{minipage}
	\hfill
	\begin{minipage}[h]{0.48\linewidth}
		\center{\includegraphics[width=\linewidth]{Decision_rule_gklshC_4} \\ b)}
	\end{minipage}
	\caption{Frequency of the use of different estimates of the Lipschitz constant for the problems of the “Simple” class (a) and of the “Hard” class (b) for $N=4$}
	\label{fig:frequency_N=4}
\end{figure}

	The dashed lines in Fig. \ref{fig:frequency_N=4} correspond to the frequency of the use of the lower estimates of the Lipschitz constant, the solid lines -- to the frequency of the use of the upper estimate. The arrangement of the curves shows the lower estimate of the Lipschitz constant to be used more frequently  at the beginning of the search process whereas at the final stage both estimates are used with almost equal frequency. These plots coincide qualitatively with the frequencies of the use of different estimates of the Lipschitz constant for other series of problems as well.

@@
